---
title: "Introduction - Illustrative example"
author: "Theophanis Tsandilas and Géry Casiez"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We detail our analysis for the illustrative example presented in the Introduction. 

We use [data](example-data.csv) from a hypothetical experiment that evaluates the time performance of three user interface techniques (*A*, *B*, and *C*) that help users complete image editing tasks of four different difficulty levels: *Level1*, *Level2*, *Level3*, and *Level4*. 

The experiment follows a fully balanced repeated-measures factorial design, where each participant ($N = 12$) performs 12 tasks in a unique order. The researcher measures the time that it takes participants to complete each task. 

### Preliminary
Let us first import all required libraries.  

```{r, echo = TRUE, message = FALSE}
library(lmerTest) # Lmer functions
library(ARTool) # ARTool library
library(effectsize) # To estimate effect sizes 

# Additional libraries for plotting and reporting
library(kableExtra) 
library(gridExtra)
library(ggplot2) 
library(tidyverse)
```

### Reading the dataset

We read our dataset and fix the type of the factors:
```{r, warning=FALSE}
df <- read.csv("example-data.csv", sep=",", header=TRUE, strip.white=TRUE)

df$Participant <- factor(df$Participant)
df$Difficulty <- factor(df$Difficulty, ordered = TRUE) # We can consider it as ordinal
df$Technique <- factor(df$Technique)
``` 

We present our dataset with a table:
```{r, warning=FALSE}
kbl(df, booktabs = TRUE) %>% kable_styling(font_size=14) %>% kable_paper(position = "center") %>% scroll_box(height = "170px")
``` 

### Plotting the data

We then plot our data to observe the main trends.

```{r, warning=FALSE, fig.height=2.7, fig.width=8}
cbPalette <- c("#999999", "#E69F00", "#F15854")

p1 <- (df %>%  group_by(Participant,Technique) %>% 
summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Technique, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Median Time (min)") +
        ylim(0, 3) + 
        geom_jitter(shape=20, position=position_jitter(0.1)) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=cbPalette)


p2 <- (df %>%  group_by(Participant,Difficulty,Technique) %>% 
           summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Difficulty, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Median Time (min)") +
        ylim(0, 6) +
        geom_point(size = 0.7, position=position_jitterdodge(0.1)) +
        theme_bw() + theme(legend.position = "none")) +
        scale_fill_manual(values=cbPalette)

grid.arrange(p1, p2, nrow = 1, widths = c(1.5, 2.5))
```

### Statistical analysis

We conduct an analysis with five different methods: PAR, LOG, RNK, ART, and INT.
```{r, warning=FALSE}
# Implementation of the Inverse Normal Transform
int = function(x){
   qnorm((rank(x) - 0.5)/length(x))
}

mpar <- lmer(Time ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - no transformation
mlog <- lmer(log(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Parametric - Log-transformation
mart <- art(Time ~ Difficulty*Technique + (1|Participant), data=df) # ARTool
mrnk <- lmer(rank(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Rank transformation 
mint <- lmer(int(Time) ~ Difficulty*Technique + (1|Participant), data=df) # Inverse Transform transformation
``` 

### Main and interaction effects
We finally report the results of our analysis:
```{r, warning=FALSE}
# Function for printing the results for each method
printAnalysis <- function(num, method, model) {
  cat(num, ". Analysis using ", method, "\n\n", sep="")

  print(anova(model))
  print(eta_squared(model))
  print(cohens_f(model))
}

printAnalysis(1, "Logarithmic Transformation", mlog)
printAnalysis(2, "ART", mart)
printAnalysis(3, "Rank Transformation", mrnk)
printAnalysis(4, "Inverse Normal Transformation", mint)
printAnalysis(5, "No Transformation", mpar)
``` 

### Contrasts
Finally, we conduct an analysis of pairwise differences between the three techniques *A*, *B*, and *C*, using a Bonferroni correction. For ART, we use the ```art.con``` function.

```{r, warning=FALSE}
library("emmeans")

# Function for printing the results for each method
printContrasts <- function(num, method, expr, model) {
  cat(num, ". Contrasts using ", method, "\n\n", sep="")

  if(method == "ART") art.con(model, expr, adjust = "bonferroni")
  else contrast(emmeans(model, expr), method="pairwise", adjust = "bonferroni")
}

printContrasts(1, "Logarithmic Transformation", ~Technique, mlog)
printContrasts(2, "ART", ~Technique, mart)
printContrasts(3, "Rank Transformation", ~Technique, mrnk)
printContrasts(4, "Inverse Normal Transformation", ~Technique, mint)
printContrasts(5, "No Transformation", ~Technique, mpar)
``` 

We observe again a significant divergence between the results of ART and the results of the other methods. 
