---
title: "Removable interactions"
author: "Theophanis Tsandilas and Géry Casiez"
#date: "2024-06-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We present here our analysis of removable interactions, where our [dataset](removable_interactions.csv) comes from a fictional experiment (within-participants design with $n = 24$) that evaluates the performance of two techniques (*Tech A* and *Tech B*) under two task difficulty levels (*easy* vs. *hard*). 

We assume that the researchers measure two dependent variables: task-completion time and perceived performance, which is measured through a five-level ordinal scale (from "very quick" to "very slow"). 

### Reading the dataset
We read our dataset:

```{r, warning=FALSE, message = FALSE}
df <- read.csv("removable_interactions.csv", sep=",", header=TRUE, strip.white=TRUE)
df$SID <- factor(df$SID)
df$Difficulty <- factor(df$Difficulty, ordered = TRUE) # We can consider it as ordinal
df$Technique <- factor(df$Technique)
``` 

We present our dataset in a table:
```{r, warning=FALSE, message = FALSE}
library(kableExtra) 
library(tidyverse)
kbl(df) %>% kable_paper(position = "center") %>% scroll_box(height = "170px")
``` 


### Analysis of task-completion time
We conduct an analysis for the Time measure, using the methods PAR, LOG, RNK, ART, and INT: 

```{r, warning=FALSE, message = FALSE}
# Implementation of the Inverse Normal Transform
library(lmerTest) # Lmer functions
library(ARTool)  # ARTool library

# Inverse normal transform
int = function(x){
   qnorm((rank(x) - 0.5)/length(x))
}

mt_par <- lmer(Time ~ Difficulty*Technique + (1|SID), data = df)
mt_rnk <- lmer(rank(Time) ~ Difficulty*Technique + (1|SID), data = df)
mt_int <- lmer(int(Time) ~ Difficulty*Technique + (1|SID), data = df)
mt_art <- art(Time ~ Difficulty*Technique + (1|SID), data = df)
mt_log <- lmer(log(Time) ~ Difficulty*Technique + (1|SID), data = df)
``` 

Let us now report the results of our analysis. Observe the very low $p$-value for interaction obtained with ART and compare it to the $p$-values obtained with the other methods.

```{r, warning=FALSE, message = FALSE}
# Function for printing the results for each method
printAnalysis <- function(num, method, model, anova = TRUE) {
  cat(num, ". Analysis using ", method, "\n\n", sep="")
  if(anova) print(anova(model))
  else print(model)
}

printAnalysis(1, "No Transformation", mt_par)
printAnalysis(2, "Logarithmic Transformation", mt_log)
printAnalysis(3, "ART", mt_art)
printAnalysis(4, "Rank Transformation", mt_rnk)
printAnalysis(5, "Inverse Normal Transformation", mt_int)
``` 

### Analysis of perceived performance
We then conduct an analysis of perceived performance using the following methods: PAR, RNK, ART, INT, and ATS (Anova-Type Statistic).

```{r, warning=FALSE, message = FALSE}
# Perceived Performance
mp_par <- lmer(PerceivedPerformance ~ Difficulty*Technique + (1|SID), data = df)
mp_rnk <- lmer(rank(PerceivedPerformance) ~ Difficulty*Technique + (1|SID), data = df)
mp_int <- lmer(int(PerceivedPerformance) ~ Difficulty*Technique + (1|SID), data = df)
mp_art <- art(PerceivedPerformance ~ Difficulty*Technique + (1|SID), data = df)

# Also try the ANOVA-type statistic (ATS)
library(nparLD) # https://www.quantargo.com/help/r/latest/packages/nparLD/2.1/nparLD
mp_ats <- nparLD(PerceivedPerformance ~ Difficulty*Technique, data=df, subject="SID", description=FALSE, order.warning = FALSE)
``` 

Let us now report the results of our analysis. We observe that all methods detect a statistically significant interaction effect ($p$-value $< .01$)

```{r, warning=FALSE, message = FALSE}

printAnalysis(1, "No Transformation", mp_par)
printAnalysis(2, "ART", mp_art)
printAnalysis(3, "Rank Transformation", mp_rnk)
printAnalysis(4, "Inverse Normal Transformation", mp_int)
printAnalysis(5, "Anova-Type Statistic (ATS)", mp_ats$ANOVA.test, anova = FALSE)
``` 

### Ordered probit models
We conduct an analysis using a ordered probit models.

We will first follow a frequentist approach, using the [```ordinal``` package](https://cran.r-project.org/web/packages/ordinal/). We assume flexible ordinal thresholds:

```{r, warning=FALSE, message = FALSE}
library(ordinal)

df$PerceivedPerformance <- factor(df$PerceivedPerformance, ordered = TRUE)
# Frequentist probit/logit model 
mp_prob <- clmm(PerceivedPerformance ~ Difficulty*Technique + (1|SID), 
                link ="probit", 
                threshold = "flexible", 
                data=df)
```

Let's print the estimates with their 95% confidence intervals, where the first four lines correspond to estimates of the thresholds:

```{r, warning=FALSE, message = FALSE}
confint(mp_prob)
``` 

 We observe that the confidence interval of the interaction term includes zero, showing no evidence for interaction. 
 
 We can also create a model with no interaction term, and then compare the two models:
 
```{r, warning=FALSE, message = FALSE}
mp_prob_2 <- clmm(PerceivedPerformance ~ Difficulty + Technique + (1|SID), 
                link ="probit", 
                threshold = "flexible", 
                data=df)

anova(mp_prob, mp_prob_2) # This is to compare the two models
``` 
 
We observe that the AIC is lower for the simpler model with no interaction, while the $\chi^2$ test provides no support for including the interaction term. 
 
We also conduct the analysis with a [Bayesian probit model](https://osf.io/preprints/psyarxiv/x8swp). We assume equal variances and do not provide any priors:

```{r, warning=FALSE, message = FALSE, results='hide'}
# We based our analysis on a Bayesian approach.
# See: https://osf.io/preprints/psyarxiv/x8swp
library(brms)

# The mo(Difficulty) formulation treats difficultly as an ordered variable
# but you could omit this formulation
 mp_brm <- brm(
   formula = PerceivedPerformance ~ Difficulty*Technique + (1|SID),
   data = df,
   family = cumulative("probit", threshold="flexible")
 )
```

If we print a summary of the results, we see again that the credible interval for the interaction term includes zero, showing no evidence for an interaction effect:
```{r, warning=FALSE, message = FALSE}
summary(mp_brm)
```


