
### Type I error rates in ratio scales {#ratio}
We first evaluate Type I error rates for ratio scales. We test a 4 $\times$ 3 within-subject design, where we refer to the 4-level factor as $X_1$ and the 3-level factor as $X_2$. We encourage readers to use the interactive tools provided in the figures to explore the trends in greater detail. 

**Main effects**. @fig-ratio-main presents Type I error rates for the main effect of $X_2$ as the magnitude of the main effect of $X_1$ increases.

::: {#fig-ratio-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Type_I_4x3_ratio"
alpha <- .05
distributions <- c("norm", "lnorm", "exp", "cauchy", "poisson", "binom")
dnames <- c("Normal", "Log-normal", "Exponential", "Cauchy", "Poisson", "Binomial")
df <- readly_data(prefix, alpha, 1, distributions, dnames)
plotly_error(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 102)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
:::

Overall, RNK and INT exhibit very good behavior across all distributions. The standard parametric ANOVA (PAR) maintains error rates below $5\%$, but for some distributions (log-normal, Cauchy, and Poisson) the error rates are noticeably lower, suggesting a loss of statistical power. Consistent with previous findings [@Mansouri:1995; @elkin:2021], ART fails to control the Type I error rate under the Cauchy distribution. However, we further show that ART is also problematic for other non-normal distributions. As the main effect of the first factor $X_1$ increases, Type I errors on the second factor $X_2$ increase dramatically, approaching $100\%$ when $a_1=8$. For the binomial distribution, error rates are already elevated ($\approx 9\%$ for $n = 20$) even when $a_1=0$, and increase further as the sample size grows.

**Contrasts**. The same issues arise when running ART's contrast procedure [@elkin:2021]. @fig-ratio-contrasts shows the corresponding results, where we report average error rates for three pairwise comparisons (since $X_2$ has three levels). In the rest of the paper, we focus on overall effects, since the results for contrasts exhibit the same patterns. 

::: {#fig-ratio-contrasts}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Type_I_4x3_ratio_contrasts"

df <- readly_data(prefix, alpha, 1, distributions, dnames)
plotly_error(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 101)
```
Average Type I error rates ($\alpha = .05$) for **contrasts on factor $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ 
:::

**Interaction effects**. @fig-ratio-interaction-1 presents Type I error rates for the $X_1 \times X_2$ interaction when $a_2=0$ and $a_1$ increases. Overall, we observe trends similar to those for main effects. ART again fails in comparable ways, although its error rates are slightly lower in this case. In addition, PAR tends to inflate error rates under the log-normal and exponential distributions as $a_1$ increases.

::: {#fig-ratio-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Type_I_4x3_ratio"
df <- readly_data(prefix, alpha, 1, distributions, dnames)

plotly_error(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 100)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
:::

The scenarios above do not capture the failure cases of the rank transformation discussed earlier (see @fig-interactions-rank), as these arise only when both coefficients $a_1$ and $a_2$ are non-zero. Because multiple definitions of interaction effects are possible in such cases, we address them in a separate subsection.

**Analysis for large samples.** We also examine Type I error rates for larger sample sizes (up to $n=512$), fixing $a_1=1$. As shown in @fig-ratio-large-samples, ART's error rates increase only slightly for log-normal data but rise substantially under the Cauchy, Poisson, and binomial distributions. These patters are consistent for both the main effect of $X_2$ and the interaction effect. As we later observe for ordinal scales, ART's tendency for increasing Type I error rates with sample size appears to be a more general issue for discrete distributions.    

::: {#fig-ratio-large-samples}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Type_I_4x3_ratio_large_samples"

distributions <- c("norm", "lnorm", "exp", "cauchy", "poisson", "binom")
dnames <- c("Normal", "Log-normal", "Exponential", "Cauchy", "Poisson", "Binomial")

df <- readly_data_1(prefix, alpha, 1, distributions, dnames)
plotly_error_by_effect_1(df, xlab = "n", max = 86)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** and the **interaction** $X_1 \times X_2$ as a function of the size $n$ of the sample. The effect on $X_1$ is non-null, where we set $a_1 = 1$.
:::

