### Type I error rates in ordinal scales {#ordinal}
We also evaluate Type I error rates for ordinal scales. We test again a $4 \times 3$ repeated-measures design. 

**Main effects**. @fig-ordinal-main present Type I errors for $X_2$'s main effect, as we vary $X_1$'s magnitude of main effect. 

::: {#fig-ordinal-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Type_I_ordinal"

distributions <- c("likert_5", "likert_5_flex", "likert_7", "likert_7_flex", "likert_11", "likert_11_flex")
dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")
df <- readly_data(prefix, alpha, 1, distributions, dnames)

plotly_error(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 60)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
:::


Our results indicate that PAR, RNK, and INT consistently maintain error rates close to $5\%$ across all tested ordinal scales. In contrast, ART exhibits a significant inflation of error rates, although this issue becomes less severe when the number of levels increases. Notably, error rates are more pronounced for flexible thresholds and tend to increase with sample size. 

**Interaction effects**. @fig-ordinal-interaction-1 displays error rates for the interaction $X_1 \times X_2$ with a single main effect applied to $X_1$. The observed patterns align with our previous findings; ART consistently inflates error rates.  

::: {#fig-ordinal-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}

plotly_error(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 50)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
:::

**Analysis for large samples.** Let us further explore ART's error rates when sample sizes become larger. As shown in @fig-ordinal-large-samples, error rates explode even when all population effects are null. The patterns are similar for all three effects.  

::: {#fig-ordinal-large-samples}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Type_I_4x3_ordinal_large_samples"

df <- readly_data_1(prefix, alpha, 0, distributions, dnames)
plotly_error_by_effect_2(df, xlab = "n", max = 100)
```
Type I error rates ($\alpha = .05$) as a function of the size $n$ of the sample, where all population effects are null.
:::
